The Role of Advection Versus Diffusion in Intraventricular Transport

### 2025-02-23 paper/main.tex
edit introduction and discussion

\keywords{cerebrospinal fluid, brain ventricles, transport, finite element method}

@@

\section{Introduction}
Brain development and function depend on numerous regulatory processes,
including interactions with cerebrospinal fluid
(CSF). CSF is a clear liquid produced by the choroid plexus
that flows through the ventricular system and subarachnoid space.
By delivering growth factors, nutrients, neurotransmitters and metabolites
and removing waste products, CSF regulates brain development and homeostasis.
Disturbances of CSF circulation can lead to brain disorders, such as
hydrocephalus where accumulation of CSF often requires emergency surgical treatment.

CSF movement occurs through multiple physiological processes,
including the action of motile cilia on the ventricular walls,
CSF secretion by the choroid plexus, and pressure gradients generated
by the cardiac cycle, respiration, and bodily movement. Motile cilia
are microscopic hair-like structures that beat in a metachronal wave
pattern to mediate fluid flow. Due to their microscopic size, ranging
from 6 to 15 microns, they contribute to flow primarily at the surface
of the brain ventricles. In contrast, pressure gradients induce fluid
flow \fixme{further from the ventricular walls (bulk flow)} and along the
perivascular space surrounding the major blood vessels.

To date, it remains poorly understood how cilia contribute to the
overall fluid flow and solute transport in the brain.
These questions have been challenging to tackle due to technical
limitations. Notably, cilia are microscopic and beat at high
frequencies ranging from 10 to 30 Hz, while other fluid dynamic
processes in the brain are slower and span larger volumes, making
cilia dynamics challenging to monitor. Until now, most studies have
analyzed one physical mechanism at a time, especially in mammals.
For instance, studies using brain explants have convincingly shown
that cilia generate complex flow patterns at the surface of the brain
ventricles; however, they failed to identify the role of pressure
gradients associated with respiration and heart pulsations in CSF
transport. Similarly, while MRI technology can measure CSF movement
in humans, it cannot detect the microscopic movements of cilia to
understand their impact in the large human brain. Consequently, we
lack a comprehensive understanding of how cilia contribute to CSF
movement and solute transport, and whether cilia dysfunction will
alter CSF composition and lead to brain disorders. This is particularly
important since ventricular enlargement, known as hydrocephalus,
is observed in animal models and humans with motile cilia deficiency,
\fixme{potentially affecting brain activity and behavior.}

The technical limitations of studying cilia-mediated flow and transport opens the
possibility of mathematical modeling and computational investigations. Numerous previous studies
focused on modeling the interaction between single or several cilia and surrounding
fluids~\cite{Guo2020SimulatingGeometries, Ruvalcaba2021NumericalTree, Smith2009MathematicalFluids,
Cui2019NumericalMethod, Cui2022AFlow}.
Others considered applications such as ciliary transport of mucous in
airways~\cite{Fulford1986Muco-ciliaryLung, Ramirez-SanJuan2020Multi-scaleArrays},
propulsion and locomotion of ciliated organisms~\cite{BLAKE1974MechanicsMotion,
Jahn1972LocomotionProtozoa, Brennen1977FluidFlagella},
coral reefs~\cite{Pacherres2022CiliaryProduction}, and the brain~\cite{Siyahhan2014FlowVentricles,
Yoshida2022EffectVentricles, Salman2022ComputationalEmbryo, Thouvenin2020OriginCanal}.
These simulation studies have provided valuable insights on how cilia mediate complex
biological flows, and how cilia may contribute to tissue physiology.

In this study, we combined computational and experimental approaches to investigate
the role of motile cilia in CSF flow and solute transport.
First, we computed flow patterns and intraventricular transport \emph{in silico}. To
this end, we used the geometry of the zebrafish larval brain ventricles,
for which we have detailed knowledge of cilia properties and CSF movement velocities
and frequencies. Notably, this early developmental stage allows us to simplify our model
by omitting CSF secretion, which starts at later stages, while still being well-conserved with mammals.
Next, we validated our computation model with experimental data by monitoring
the dynamics of a photoconvertible fluorescent protein Dendra2 expressed within the brain ventricles.
Our results show that diffusion plays a major role in the transport of small solutes,
whereas cilia majorly contribute to the movement of larger particles.
We also found that the location of cilia, as well as the ventricular geometry, greatly
impact solute distribution. Altogether, our work presents a computational
framework together with new concepts of how cilia
collaborate with other factors to distribute molecules and
particles within the brain ventricles.

@@

\subsubsection{Flow induced by cardiac pulsations}
To drive a cardiac-induced pulsatile flow inside the brain ventricles, we set $\Tilde{p}(t)=0$ and
$\Tilde{p}(t)=-A\sin{\omega t}$ on anterior and posterior parts, respectively, of the
boundary $\Gp$~(\Crefpanels{fig:fig1}{d}, blue markers).
Based on a measured cardiac frequency of $f=2.22$ Hz~\cite{Olstad2019CiliaryDevelopment},
the angular cardiac frequency $\omega=2\pi f=6.97$ rad/s.
The amplitude $A=1.5$ mPa determines the magnitude of the pulsatile flow,
and was calibrated based on experimental data~\cite{Olstad2019CiliaryDevelopment}.
Similar to $\tau$, the amplitude $A$ was chosen by comparing experimental and simulated velocities.
Particle tracking data of pulsatile motion was used to determine the contributions of the
cardiac pulsations to the velocity magnitude,
as compared to the effect of cilia motion on the flow (see~\Cref{sec:appC_particle_transport}).

@@

\subsection{Estimation of diffusion coefficients via the Stokes-Einstein relation}
Motivated by an interest in molecules that are significant in neural development,
we study the transport of molecules resembling the Dendra2 Fluorescent Protein (Dendra2)
used in this study, the Starmaker+Green Fluorescent Protein (STM-GFP) reported in
Jeong~\emph{et al.}~\cite{Jeong2024TheZebrafish}, and exosomes.
To estimate the diffusion coefficients of these, when not
available in the literature, we use the Stokes--Einstein relation
\begin{equation}
    D = \frac{k_B T}{6\pi \mu R},
    \label{eq:D_stokes_einstein}
\end{equation}
which is an expression for the diffusion coefficient of a spherical
molecule suspended in a viscous fluid~\cite{Einstein1905UberTeilchen}.
In~\Cref{eq:D_stokes_einstein}, $k_B = \num{1.38e-23}$ J/K is the Boltzmann constant,
$T = 310$ K is the absolute temperature, $\mu$ is the dynamic viscosity of the
solvent (in this case the CSF), and $R$ is the radius of the molecule.
For the exosomes, we calculate $D$ using~\Cref{eq:D_stokes_einstein}
with $R=200$ nm. This radius covers the slowest diffusing exosomes,
a choice made because smaller exosomes would have diffusion coefficients
similar to the other molecules.
For STM-GFP, we estimated the diffusion coefficient by extrapolating
the experimentally observed diffusion coefficient of
GFP~\cite{Swaminathan1997PhotobleachingDiffusion, Potma2001ReducedCells},
assuming that if $D$ scales with $1/R$ according to~\Cref{eq:D_stokes_einstein},
$D$ would scale with the cube root of the molecule mass~\cite{Goodhill1997DiffusionGuidance}.
The resulting diffusion coefficients used in the simulations
are $D_1$ (exosomes), $D_2$ (STM-GFP) and $D_3$ (Dendra2)~(\Cref{tab:diff_coeff_table}).

@@

\begin{table}[h!]
    \centering
    \caption{Molecular mass and diffusion coefficients of simulated solutes.
    STM and GFP denote Starmaker Protein and Green Fluorescent Protein, respectively.
    (*) The value was extrapolated based on the value of $D$ for
    GFP reported in the literature~\cite{Swaminathan1997PhotobleachingDiffusion, Potma2001ReducedCells},
    assuming that $D$ scales with the cube root of the molecule mass.
    (\textdagger) The value was estimated using the Stokes-Einstein relation~\Cref{eq:D_stokes_einstein}.}
    \label{tab:diff_coeff_table}
    \begin{tabular}{l|ccc}
        \toprule
        Molecule & Mass [kDa] & $D \ [\mathrm{m^2/s}]$ & Reference \\ 
        \midrule 
        Dendra2 Fluorescent Protein & 25.6~\cite{Gurskaya2006EngineeringLight}
        & $\num{1.15e-10}$  & \cite{GuraSadovsky2017MeasurementExpansion}\\
        GFP (saline aqueous solution) & 26.9~\cite{UniProt2024GFPUniProtKB}
        & $\num{8.70e-11}$ & \cite{Swaminathan1997PhotobleachingDiffusion, Potma2001ReducedCells} \\
        STM-GFP & 66.2~\cite{UniProt2024StmUniProtKB} + 26.9~\cite{UniProt2024GFPUniProtKB}
        & $\num{5.75e-11}$ & * \\
        Exosomes (radius of 200 nm) & -- & $\num{1.63e-12}$ & \textdagger \\
        \bottomrule
    \end{tabular}
\end{table}

@@

To evaluate the relative importance of advection and
diffusion as transport mechanisms, we consider the Péclet number
\begin{equation*}
    \mathrm{Pe} = \frac{\mathrm{diffusion \ timescale}}{\mathrm{advection \ timescale}}
    = \frac{t_d}{t_a} = \frac{L_p^2/D}{L_p/U_p} = \frac{L_p U_p}{D}.
\end{equation*}
Thus, if $\mathrm{Pe} > 1$, transport is dominated by advection,
and if $\mathrm{Pe} < 1$ transport is dominated by diffusion.
Here, $L_p$ and $U_p$ are characteristic length and velocity scales,
respectively, and $D$ is the solute diffusion coefficient.
In calculating a global Péclet number, we use $L_p=600$ \textmu m, the approximate length of the
ventricles geometry along the rostrocaudal ($x$-) axis.
We use a mean velocity $\overline{U}$ as the characteristic velocity $U_p$.
The mean velocity $\overline{U}$ is calculated by first averaging the velocity in space:
\begin{equation*}
    \overline{u}^2 = \frac{1}{|\Omega|}\int_{\Omega}\uu\cdot\uu\dx,
\end{equation*}
and then averaging $\overline{u}$ in time over one cardiac cycle.
To assess the assumption that the CSF flow is governed by the Stokes equations,
we also compute the Reynolds number
\begin{equation*}
    \mathrm{Re} = \frac{\mathrm{inertial \ forces}}{\mathrm{viscous \ forces}} = \frac{\rho U_r L_r}{\mu},
\end{equation*}
with $L_r$ and $U_r$ as length and velocity scales characteristic of the flow.

@@

To quantify the transport of solute, we plot the profile of the mean
concentrations $\cbar_i(t)$ as functions of time
(\Crefpanels{fig:fig2}{e--g}) and
the times $\hat{t}$ when the mean concentrations $\cbar_i$ first exceed
a threshold value $\hat{c}$ (\Crefpanels{fig:fig2}{h})
in 6 ROIs (\Crefpanels{fig:fig2}{a}).
The dynamics at the site of photoconversion (ROI 1)
is logarithmic and follows from imposing \eqref{eq:photoconversion_curve}.
For ROIs 2--6, we observe faster kinetics with higher diffusion coefficients
(\Crefpanels{fig:fig2}{e--g}),
which results in smaller times-to-threshold
(\Crefpanels{fig:fig2}{h}).
For all three diffusion coefficients, the dynamics of ROI 5
is significantly delayed compared to ROI 2--4,
and ROI 6 is the slowest. Notably, for $D_1$ the mean concentration
in ROI 6 is close to zero at the end of simulation.
To quantify the balance of advection versus diffusion as transport mechanisms,
we calculate the Péclet numbers $\mathrm{Pe}_1=519$,
$\mathrm{Pe}_2=14.7$ and $\mathrm{Pe}_3=7.4$ for $D_1$, $D_2$ and $D_3$, respectively.
Since $\mathrm{Pe}>1$, this indicates that advection is the
dominant transport mechanism on a global scale.

@@

Meanwhile, we identified that transport of larger molecules
(thus with smaller diffusion coefficients),
such as 200 nm exosomes, are almost entirely dependent on advection.
Indeed, there was essentially zero solute in the posterior ventricle at the
final simulation time (around 850 s), a ventricle where transport is largely driven
by diffusion due to the absence of motile cilia.
For $D_1$, an estimate of the diffusion timescale $t_d=L^2/D_1$ over a distance
$L=200$ \textmu m from the middle to the posterior ventricle is approximately 24500 s.
Advective transport by the mean velocity $\overline{U}=1.41 \ \mu$m/s over the same
distance would happen on a timescale of the order of $t_a=L/\overline{U}=141$ s.
Comparing this with the minuscule values of $\cbar_6$ after simulating transport
with $D_1$ for 850 s, one can conclude that if advection
played a role in transporting solute from the photoconversion site in the middle ventricle
to the posterior ventricle, values of $\cbar_6$ would be higher than those observed.

\subsection{Impact of cilia motility and ventricular morphology}
To gain further insights into how CSF flow and intraventricular transport depend on cilia
motility and ventricular morphology, we manipulated certain
parameters in our models. First, we paralyzed cilia in specific regions of
our model. We identified that the cilia in the dorsal middle ventricle contribute
the most to solute distribution, which aligns well with the fact
that these cilia generate the strongest flow field in vivo~\cite{Olstad2019CiliaryDevelopment}
and in our model. In contrast, removing the ventral cilia in the middle ventricle
did not greatly impact flow dynamics, despite these cilia contributing a third of
the total force exerted by the cilia. However, removing ventral cilia in the middle ventricle
lead to reduced transport towards the posterior ventricle,
meanwhile increasing transport towards the anterior ventricle,
as the ventral cilia help induce flow towards the posterior ventricle.
Removing the cilia in the anterior ventricle did not affect the transport.

Next, we modified the ventricular geometry and identified the impact on transport.
Notably, shrinking ducts reduced interventricular solute movement, while shrinking
the middle ventricle reduced diffusive transport upstream of the flow field.
We previously identified large inter-individual variation in ventricular size and
CSF velocities between larva at given developmental stages and throughout
development~\cite{Olstad2019CiliaryDevelopment, DGama2025MotileBrain, Jeong2024TheZebrafish}.
Our results suggest that these two parameters, CSF velocities and ventricular size,
may well co-vary and explain the variation in the two.

@@

\lyng{HH: Compare numerical results with transport results in Figure 7 in \cite{Olstad2019CiliaryDevelopment}.}

\lyng{HH: Discuss cilia serving as an "enhanced mixing factor" on small scales,
refer to this~\cite{Siyahhan2014FlowVentricles} and this~\cite{Yoshida2022EffectVentricles}
paper that studied this in human brain ventricles.
Include a discussion of what happened when these papers removed cilia.}

@@

A priori error estimates for the $\mathrm{BDM}_k$-$\mathrm{DG}_{k-1}$ scheme are naturally 
derived in a mesh-dependent broken $H^1$-norm for $\uu$ and the $L^2$-norm for the pressure. The respective 
errors are expected to decay with order $k$,
see for example Cockburn~\emph{et al.}~\cite{Cockburn2005AEquations}.
For the sake of simplicity,~\Cref{tab:cylinder_error_rates}
reports the velocity errors in the standard $H^1$ norm. Using polynomial degree  
$k=1$, we observe linear convergence in both the velocity and the pressure.   
\mk{MK: I see 1.5 in the table for $\uu$ in $L^2$?}
\lyng{HH: Yes, while the $H^1$ error decays linearly, isn't this in line with expectations?
Could be more specific: "Using polynomial degree 
$k=1$, we observe linear convergence in both the velocity $H^1$ error and the pressure $L^2$ error."}

@@

\medskip
\textbf{Data Availability Statement} \par
\fixme{TO DO}.

\medskip
\textbf{Declaration of No Competing Interest} \par
\fixme{TO DO}.

@@

% Acknowledgements
\medskip
\textbf{Acknowledgements} \par 
Halvor Herlyng acknowledges support from the national infrastructure for computational
science in Norway, Sigma2, via grant \#NN8049K.
This work was funded by the Research Council of Norway (NJY FRIPRO grants 314189 and 353348).
Thanks are extended to J\o rgen Dokken for valuable model implementation discussions.

@@

\textbf{Only-cilia/no-cardiac model.} This model version has cilia forces
as the only mechanism driving flow. We consider the
Stokes equations~\eqref{eq:appendix_stokes_eqs} with the slip boundary conditions
\begin{subequations}
    \begin{alignat}{2}
        \uu\cdot\nn &= 0 \quad &&\mathrm{on} \ \Gamma=\Gc\cup\Gs, \label{eq:modelA_slip_BC_impermeability} \\
        \bsigpar &= \btau \quad &&\mathrm{on} \ \Gc,
        \label{eq:modelA_slip_BC_shear} \\
        \bsigpar &= \mathbf{0} \quad &&\mathrm{on} \ \Gs,\label{eq:modelA_full_slip_BC}
    \end{alignat}\label{eq:modelA_slip_BC}%
\end{subequations}%
on the boundary $\Gamma=\Gamma_c\cup\Gamma_s$. The boundary $\Gamma_s$ is a free-slip boundary,
where we impose the free-slip condition~\eqref{eq:modelA_full_slip_BC}.
The first condition~\eqref{eq:modelA_slip_BC_impermeability} is the
impermeability condition, enforcing no flow normal to the boundary,
whereas~\eqref{eq:modelA_slip_BC_shear} imposes a stress tangential to
the boundary. These tangential stresses are applied to represent
the forces that the collective cilia motion exerts on the cerebrospinal fluid.

The boundary conditions~\eqref{eq:modelA_slip_BC} are set weakly through the stress boundary integral 

@@

in the weak form~\eqref{eq:modelA_weak_form1} of the Stokes equations.

>>>>

\keywords{cerebrospinal fluid, brain ventricles, transport, finite element methods}

@@

\section{Introduction}
Brain development and function depend on numerous regulatory processes,
including the flow of cerebrospinal fluid
(CSF). CSF is a clear liquid produced by the choroid plexus
and the fluid flows through the ventricular system and subarachnoid space.
By transporting growth factors, nutrients, neurotransmitters and metabolites
and removing waste products, CSF regulates brain development and homeostasis.
Disturbance of the CSF circulation can lead to brain disorders, such as
hydrocephalus where accumulation of CSF often requires emergency surgical treatment.

Motion of CSF occurs through multiple physiological processes,
including the beating motion of motile cilia lining the ventricular walls,
CSF secretion by the choroid plexus, and pressure gradients generated
by the cardiac cycle, respiration, and bodily movement. Motile cilia
are microscopic hair-like structures that beat in a metachronal wave
pattern to mediate fluid flow. Due to their microscopic size, ranging
from 6 to 15 microns, the cilia contribute to flow primarily at the surface
of the brain ventricles. In contrast, fluid flow induced by the 
cardiac and respiratory cycles account for the bulk flow in the brain
ventricles~\cite{Vinje2019RespiratoryMeasurements},
as well as the CSF flow in the perivascular spaces.

To date, it remains poorly understood how cilia contribute to the
overall fluid flow and solute transport within the brain.
These questions have been challenging to tackle due to technical
limitations. Notably, cilia are microscopic and beat at high
frequencies ranging from 10 to 30 Hz, while other fluid dynamic
processes in the brain are slower and span larger volumes, making
cilia dynamics challenging to monitor in vivo. Until now, most studies have
analyzed one physical mechanism at a time, especially in mammals.
For instance, studies using brain explants have convincingly shown
that cilia generate complex flow patterns at the surface of the brain
ventricles; however, they failed to identify the role of pressure
gradients associated with respiration and cardiac pulsations in CSF flow
and the accompanying transport. Similarly, while MRI technology can measure CSF movement
in humans, it cannot detect the microscopic movements of cilia to
understand their impact in the large human brain. Consequently, we
lack a comprehensive understanding of how cilia contribute to CSF
motion and solute transport, and whether cilia dysfunction will
alter CSF composition and lead to brain disorders. This is particularly
important since ventricular enlargement, known as hydrocephalus,
is observed in animal models and humans with motile cilia deficiency,
indiciating that the cilia play a crucial role for healthy brain functioning.

The technical limitations of studying cilia-mediated flow and transport opens the
possibility of mathematical modeling and computational investigations. Numerous previous studies
have focused on modeling the interaction between single or several cilia and surrounding
fluids~\cite{Guo2020SimulatingGeometries, Ruvalcaba2021NumericalTree, Smith2009MathematicalFluids,
Cui2019NumericalMethod, Cui2022AFlow}.
Others considered applications such as ciliary transport of mucous and particles in
airways~\cite{Fulford1986Muco-ciliaryLung, Ramirez-SanJuan2020Multi-scaleArrays},
propulsion and locomotion of ciliated organisms~\cite{BLAKE1974MechanicsMotion,
Jahn1972LocomotionProtozoa, Brennen1977FluidFlagella},
oxygen regulation in coral reefs~\cite{Pacherres2022CiliaryProduction},
and cilia-mediated flow in the brain~\cite{Siyahhan2014FlowVentricles,
Yoshida2022EffectVentricles, Salman2022ComputationalEmbryo, Thouvenin2020OriginCanal}.
These simulation studies have provided valuable insights on how cilia mediate complex
biological flows, and how cilia may contribute to tissue physiology.

In this study, we combined computational and experimental approaches to investigate
the role of motile cilia in intraventricular CSF flow and transport.
First, we computed flow patterns and simulated solute transport with a computational model. To
this end, we used the geometry of the zebrafish larval brain ventricles,
for which we have detailed knowledge of cilia properties and CSF motion velocities
and frequencies. Notably, this early developmental stage allows us to simplify our model
by omitting CSF secretion, which starts at later stages,
while still being well-conserved with mammals. 
Next, we validated our computation model with experimental data by monitoring in vivo
the dynamics of a photoconvertible fluorescent protein Dendra2
expressed within the brain ventricles.
Our results show that diffusion plays a major role in the transport of small solutes,
whereas cilia majorly contribute to the movement of larger particles.
We also found that the location of cilia, as well as the ventricular geometry, greatly
impact solute distribution. Altogether, our work presents a computational
framework together with new concepts of how cilia contribute to distribution of molecules and
particles within the brain ventricles.

@@

\subsubsection{Flow induced by cardiac pulsations}
To generate a cardiac-induced pulsatile flow inside the brain ventricles, we set $\Tilde{p}(t)=0$ and
$\Tilde{p}(t)=-A\sin{\omega t}$ on anterior and posterior parts, respectively, of the
boundary $\Gp$~(\Crefpanels{fig:fig1}{d}, blue markers).
Based on a measured cardiac frequency of $f=2.22$ Hz~\cite{Olstad2019CiliaryDevelopment},
the angular cardiac frequency $\omega=2\pi f=6.97$ rad/s.
The amplitude $A=1.5$ mPa determines the magnitude of the pulsatile flow,
and was calibrated based on experimental data~\cite{Olstad2019CiliaryDevelopment}.
Similar to $\tau$, the amplitude $A$ was chosen by comparing experimental and simulated velocities.
Particle tracking data of pulsatile motion was used to determine the contributions of the
cardiac pulsations to the velocity magnitude,
as compared to the effect of cilia motion on the flow (cf.~\Cref{sec:appC_particle_transport}).

@@

\subsection{Estimation of diffusion coefficients via the Stokes-Einstein relation}
Motivated by an interest in molecules that are significant in neural development,
we study the transport of molecules resembling the Dendra2 Fluorescent Protein (Dendra2)
used in this study, the Starmaker+Green Fluorescent Protein (STM-GFP) reported in
Jeong~\emph{et al.}~\cite{Jeong2024TheZebrafish}, and exosomes.
To estimate the diffusion coefficients of these, when not
available in the literature, we use the Stokes--Einstein relation
\begin{equation}
    D = \frac{k_B T}{6\pi \mu R},
    \label{eq:D_stokes_einstein}
\end{equation}
which is an expression for the diffusion coefficient of a spherical
molecule suspended in a viscous fluid~\cite{Einstein1905UberTeilchen}.
In~\Cref{eq:D_stokes_einstein}, $k_B = \num{1.38e-23}$ J/K is the Boltzmann constant,
$T = 310$ K is the absolute temperature, $\mu$ is the dynamic viscosity of the
solvent (in this case the CSF), and $R$ is the radius of the molecule.
For the exosomes, we calculate $D$ using~\Cref{eq:D_stokes_einstein}
with $R=150$ nm, which represents exosomes at the larger end of the
scale~\cite{Moghassemi2024ExtracellularDecade}.
Hence this radius covers the slowest diffusing exosomes,
a choice made because smaller exosomes would have diffusion coefficients
similar to the other molecules we consider.
For STM-GFP, we estimated the diffusion coefficient by extrapolating
the experimentally observed diffusion coefficient of
GFP~\cite{Swaminathan1997PhotobleachingDiffusion, Potma2001ReducedCells},
assuming that if $D$ scales with $1/R$ according to~\Cref{eq:D_stokes_einstein},
$D$ would scale with the cube root of the molecule mass~\cite{Goodhill1997DiffusionGuidance}.
The resulting diffusion coefficients used in the simulations
are $D_1$ (exosomes), $D_2$ (STM-GFP) and $D_3$ (Dendra2)~(\Cref{tab:diff_coeff_table}).

@@

\begin{table}[h!]
    \centering
    \caption{Molecular mass and diffusion coefficients of simulated solutes.
    STM and GFP denote Starmaker Protein and Green Fluorescent Protein, respectively.
    (*) The value was extrapolated based on the value of $D$ for
    GFP reported in the literature~\cite{Swaminathan1997PhotobleachingDiffusion, Potma2001ReducedCells},
    assuming that $D$ scales with the cube root of the molecule mass.
    (\textdagger) The value was estimated using the Stokes-Einstein relation~\Cref{eq:D_stokes_einstein}.}
    \label{tab:diff_coeff_table}
    \begin{tabular}{l|ccc}
        \toprule
        Molecule & Mass [kDa] & $D \ [\mathrm{m^2/s}]$ & Reference \\ 
        \midrule 
        Dendra2 Fluorescent Protein & 25.6~\cite{Gurskaya2006EngineeringLight}
        & $\num{1.15e-10}$  & \cite{GuraSadovsky2017MeasurementExpansion}\\
        GFP (saline aqueous solution) & 26.9~\cite{UniProt2024GFPUniProtKB}
        & $\num{8.70e-11}$ & \cite{Swaminathan1997PhotobleachingDiffusion, Potma2001ReducedCells} \\
        STM-GFP & 66.2~\cite{UniProt2024StmUniProtKB} + 26.9~\cite{UniProt2024GFPUniProtKB}
        & $\num{5.75e-11}$ & * \\
        Exosomes (radius of 150 nm) & -- & $\num{2.17e-12}$ & \textdagger \\
        \bottomrule
    \end{tabular}
\end{table}

@@

To evaluate the relative importance of advection and
diffusion as transport mechanisms, we consider the Péclet number
\begin{equation*}
    \mathrm{Pe} = \frac{\mathrm{diffusion \ timescale}}{\mathrm{advection \ timescale}}
    = \frac{t_d}{t_a} = \frac{L_p^2/D}{L_p/U_p} = \frac{L_p U_p}{D}.
\end{equation*}
Thus, if $\mathrm{Pe} > 1$, transport is dominated by advection,
and if $\mathrm{Pe} < 1$ transport is dominated by diffusion.
Here, $L_p$ and $U_p$ are characteristic length and velocity scales,
respectively, and $D$ is the solute diffusion coefficient.
In calculating a global Péclet number, we use $L_p=600$ \textmu m, the approximate length of the
ventricles geometry along the rostrocaudal ($x$-) axis.
We use a mean velocity $\overline{U}$ as the characteristic velocity $U_p$.
The mean velocity $\overline{U}$ is calculated by first averaging the velocity in space:
\begin{equation*}
    \overline{u}^2 = \frac{1}{|\Omega|}\int_{\Omega}\uu\cdot\uu\dx,
\end{equation*}
and then averaging $\overline{u}$ in time over one cardiac cycle.
To assess the assumption that inertial forces are negligible compared to
viscous forces in the CSF flow, such that the flow is governed by the Stokes equations,
we also compute the Reynolds number
\begin{equation*}
    \mathrm{Re} = \frac{\mathrm{inertial \ forces}}{\mathrm{viscous \ forces}} = \frac{\rho U_r L_r}{\mu},
\end{equation*}
with $L_r$ and $U_r$ as length and velocity scales characteristic of the flow.

@@

To quantify the transport of solute, we plot the profile of the mean
concentrations $\cbar_i(t)$ as functions of time
(\Crefpanels{fig:fig2}{e--g}) and
the times $\hat{t}$ when the mean concentrations $\cbar_i$ first exceed
a threshold value $\hat{c}$ (\Crefpanels{fig:fig2}{h})
in 6 ROIs (\Crefpanels{fig:fig2}{a}).
The dynamics at the site of photoconversion (ROI 1)
is logarithmic and follows from imposing \Cref{eq:photoconversion_curve}.
For ROIs 2--6, we observe faster kinetics with higher diffusion coefficients
(\Crefpanels{fig:fig2}{e--g}),
which results in smaller times-to-threshold
(\Crefpanels{fig:fig2}{h}).
For all three diffusion coefficients, the dynamics of ROI 5
is significantly delayed compared to ROI 2--4,
and ROI 6 is the slowest. Notably, for $D_1$ the mean concentration
in ROI 6 is close to zero at the end of simulation.
To quantify the balance of advection versus diffusion as transport mechanisms,
we calculate the Péclet numbers $\mathrm{Pe}_1=519$,
$\mathrm{Pe}_2=14.7$ and $\mathrm{Pe}_3=7.4$ for $D_1$, $D_2$ and $D_3$, respectively.
Since $\mathrm{Pe}>1$, this indicates that advection is the
dominant transport mechanism on a global scale.

@@

Meanwhile, we identified that transport of larger molecules
(thus with smaller diffusion coefficients),
such as 150 nm exosomes, are almost entirely dependent on advection.
Indeed, there was essentially zero solute in the posterior ventricle at the
final simulation time (around 850 s), a ventricle where transport is largely driven
by diffusion due to the absence of motile cilia.
For $D_1$, an estimate of the diffusion timescale $t_d=L^2/D_1$ over a distance
$L=200$ \textmu m from the middle to the posterior ventricle is approximately 24500 s.
Advective transport by the mean velocity $\overline{U}=1.41 \ \mu$m/s over the same
distance would happen on a timescale of the order of $t_a=L/\overline{U}=141$ s.
Comparing this with the minuscule values of $\cbar_6$ after simulating transport
with $D_1$ for 850 s, one can conclude that if advection
played a role in transporting solute from the photoconversion site in the middle ventricle
to the posterior ventricle, values of $\cbar_6$ would be higher than those observed.

\subsection{Effects of cilia motility and ventricular morphology}
To gain further insights into how CSF flow and intraventricular transport depend on cilia
motility and ventricular morphology, we manipulated certain
parameters in our models. First, we investigated how complete loss of cilia motility
impacted the transport dynamics. Simulations with a cardiac-only/no-cilia lead to
slower transport when compared to the baseline model that included cilia. Especially
the middle-ventricular region downstream of CSF flow (ROI 2) and the anterior ventricle (ROI 5)
experienced substantially increased times-to-threshold. We also observed
slower dynamics in the rest of the middle ventricle (ROIs 3 and 4), whereas the
posterior ventricle (ROI 6) was unaffected by the loss
of cilia motility, indicating that transport from the middle to the posterior ventricle is
purely driven by diffusion. We also performed photoconversion experiments on cilia mutant
zebrafish larva (\emph{smh}), observing statistically signifcant slower transport in 
ROIs 2 and 4. The results provide not only confidence in that our computational model
well represents the physics, but also valuable quantification of the impact of cilia motility
on intraventricular transport. Interestingly, a previous simulation study of reduced cilia motility
also observed a drastic reduction in CSF mixing in specific regions of the brain
ventricles~\cite{Yoshida2022EffectVentricles}, similar to our results. Since impaired brain 
clearance of tracers has been demonstrated in idiopathic normal pressure hydrocephalus
patients~\cite{Eide2020MagneticHydrocephalus}, cilia-enhanced mixing in the brain might be
fundamental to physiological transport within the brain.

Next, we paralyzed cilia in specific regions of
our model. We identified that the cilia in the dorsal middle ventricle contribute
the most to solute distribution, which aligns well with the fact
that these cilia generate the strongest flow field in vivo~\cite{Olstad2019CiliaryDevelopment}
and in our model. In contrast, removing the ventral cilia in the middle ventricle
did not greatly impact flow dynamics, despite these cilia contributing a third of
the total force exerted by the cilia. However, removing ventral cilia in the middle ventricle
lead to reduced transport towards the posterior ventricle,
meanwhile increasing transport towards the anterior ventricle,
as the ventral cilia help induce flow towards the posterior ventricle.
Removing the cilia in the anterior ventricle did not affect the transport.

With an interest in assessing the impact of ventricular morphology on solute transprot,
we simulated transport on modified ventricular geometries.
Notably, shrinking ducts reduced interventricular solute movement, while shrinking
the middle ventricle reduced diffusive transport upstream of the flow field.
We previously identified large inter-individual variation in ventricular size and
CSF velocities between larva at given developmental stages and throughout
development~\cite{Olstad2019CiliaryDevelopment, DGama2025MotileBrain, Jeong2024TheZebrafish}.
Our results suggest that these two parameters, CSF velocities and ventricular size,
may well co-vary and explain the variation in the two.

@@

@@

A priori error estimates for the $\mathrm{BDM}_k$-$\mathrm{DG}_{k-1}$ scheme are naturally 
derived in a mesh-dependent broken $H^k$-norm for $\uu$ and the $L^2$-norm for the pressure.
The respective error norms are expected to decay with order $k$~\cite{Cockburn2005AEquations}.
For the sake of simplicity,~\Cref{tab:cylinder_error_rates}
reports the velocity errors in the standard $H^k$ norm. Using polynomial degree  
$k=1$, we observe linear convergence in both the velocity and the pressure.   
\mk{MK: I see 1.5 in the table for $\uu$ in $L^2$?}
\lyng{HH: Yes, we should expect 2 there? Maybe be more specific in the comment:
"Using polynomial degree $k=1$, we observe linear convergence in both the velocity $H^1$ error
and the pressure $L^2$ error. For the $L^2$ error of the velocity, we observe a suboptimal
convergence order of 1.5"}

@@

\medskip
\textbf{Data Availability Statement} \par
The data and implementations presented in this paper are licensed under \fixme{LICENSE}
and openly avalaible at \fixme{REPO-LINK}.

\medskip
\textbf{Conflict of Interest Disclosure} \par
The authors declare no conflicts of interest.

@@

% Acknowledgements
\medskip
\textbf{Acknowledgements} \par 
Thanks are extended to J\o rgen Dokken for valuable model implementation discussions.
Halvor Herlyng acknowledges support from the national infrastructure for computational
science in Norway, Sigma2, via grant \#NN8049K.
This work was funded by the Research Council of Norway (NJY FRIPRO grants 314189 and 353348).

@@

\textbf{Only-cilia/no-cardiac model.} This model version has cilia forces
as the only mechanism driving flow. We consider the
Stokes equations~\Cref{eq:appendix_stokes_eqs} with the slip boundary conditions
\begin{subequations}
    \begin{alignat}{2}
        \uu\cdot\nn &= 0 \quad &&\mathrm{on} \ \Gamma=\Gc\cup\Gs,
        \label{eq:modelA_slip_BC_impermeability} \\
        \bsigpar &= \btau \quad &&\mathrm{on} \ \Gc,
        \label{eq:modelA_slip_BC_shear} \\
        \bsigpar &= \mathbf{0} \quad &&\mathrm{on} \ \Gs,\label{eq:modelA_full_slip_BC}
    \end{alignat}\label{eq:modelA_slip_BC}%
\end{subequations}%
on the boundary $\Gamma=\Gamma_c\cup\Gamma_s$. The boundary $\Gamma_s$ is a free-slip boundary,
where we impose the free-slip condition~\Cref{eq:modelA_full_slip_BC}.
The first condition~\Cref{eq:modelA_slip_BC_impermeability} is the
impermeability condition, enforcing no flow normal to the boundary,
whereas~\Cref{eq:modelA_slip_BC_shear} imposes a stress tangential to
the boundary. These tangential stresses are applied to represent
the forces that the collective cilia motion exerts on the cerebrospinal fluid.

The boundary conditions~\Cref{eq:modelA_slip_BC} are set weakly through the stress
boundary integral 

@@

in the weak form~\Cref{eq:modelA_weak_form1} of the Stokes equations.
